### 2023-02-27 abstract.tex
Updates from Overleaf

\begin{abstract}
Type systems typically only define the conditions under which an expression is well-typed, leaving ill-typed expressions entirely meaningless. 
This approach is insufficient as the basis for language servers for modern programming environments, which are expected to sensibly localize type errors and recover, i.e. continue to provide a variety of semantic services in their presence. 
This paper addresses this problem, contributing the first comprehensive formal account of type error localization and recovery, called the marked lambda calculus. This calculus, which is intended as a core calculus for language servers, consists of a gradual type system for expressions with marked errors and a total marking procedure on unmarked expressions with holes. We mechanize the metatheory of the marked lambda calculus in Agda, deploy its total marking procedure to fix several instances of undefined behavior in the Hazelnut structure editor calculus, and implement the system as the new basis for Hazel, a full-scale live functional programming environment.

The marked lambda calculus supports local type inference, i.e. it is bidirectionally typed. Constraint-based type inference schemes (e.g. Hindley-Milner) are notorious for complicating error localization. We approach the problem of incorporating constraint-based type inference by deploying constraint solving as a type hole filling layer atop a gradual bidirectionally typed core. When the system of constraints cannot be fully solved, the unification algorithm we develop localizes errors arising from inconsistent constraints to individual type holes, rather than attempting to localize them in an \emph{ad hoc} manner to one particular expression and then failing to recover from there. The user can resolve these errors by selecting from suggested partial solutions, or leave the errors unresolved, in which case the type hole recovers gradually. We formally specify this type hole inference system and implement it in Hazel. 
\end{abstract}

>>>>

\begin{abstract}
Type systems typically only define the conditions under which an expression is well-typed, leaving ill-typed expressions entirely meaningless. 
This approach is insufficient as the basis for language servers for modern programming environments, which are expected to sensibly localize any type errors and recover, i.e. continue to provide a variety of downstream semantic services. 
This paper addresses this problem, contributing the first comprehensive formal account of type error localization and recovery: the marked lambda calculus. In particular, we define a gradual type system for expressions with marked errors and holes, together with a total procedure for marking unmarked expressions. We mechanize the metatheory of the marked lambda calculus in Agda and implement it, scaled up, as the new basis for Hazel, a full-scale live functional programming environment. 
As a secondary contribution, we show how total marking resolves 
the problem of undefined behavior in the Hazelnut structure editor calculus
(though marking itself does not require a structure editor).

The marked lambda calculus supports local type inference, i.e. it is bidirectionally typed. Constraint-based type inference notoriously complicates error localization and recovery. We approach this problem by deploying constraint solving as a type-hole-filling layer atop our gradual bidirectionally typed core. The unification algorithm we contribute localizes errors arising from inconsistent constraints to type holes with traced  provenances, rather than attempting to localize them in an \emph{ad hoc} manner to particular expressions. The user can choose how to shift these errors to expressions by selecting from suggested partial solutions, or allow the unresolved type holes to recover gradually. We implement this type hole inference system in Hazel. 
\end{abstract}
